**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**

### 2021-08-30 exploratory/exploratory.Rmd
draft to send to coauthors

library(tidyverse)
library(knitr)
library(kableExtra)
library(ggpubr)
library(nationalparkcolors)
library(data.table)
library(cowplot)
library(pROC)

#given a number, adds correct commas
set_commas <- function(num){

}
```

@@

for( m in models ){
  phylum <- read_csv(paste0("../data/process/combined-phylum-",m,".csv"),
                     col_types = c(method = col_character(),.default=col_double())) %>% 
    mutate(level="phylum")
  class <- read_csv(paste0("../data/process/combined-class-",m,".csv"),
                     col_types = c(method = col_character(),.default=col_double())) %>% 
    mutate(level="class")
  order <- read_csv(paste0("../data/process/combined-order-",m,".csv"),
                     col_types = c(method = col_character(),.default=col_double())) %>% 
    mutate(level="order")
  family <- read_csv(paste0("../data/process/combined-family-",m,".csv"),
                     col_types = c(method = col_character(),.default=col_double())) %>% 
    mutate(level="family")
  genus <- read_csv(paste0("../data/process/combined-genus-",m,".csv"),
                     col_types = c(method = col_character(),.default=col_double())) %>%
    mutate(level="genus")
  otu <- read_csv(paste0("../data/process/combined-otu-",m,".csv"),
                     col_types = c(method = col_character(),.default=col_double())) %>%
    mutate(level="otu")
  asv <- read_csv(paste0("../data/process/combined-asv-",m,".csv"),
                     col_types = c(method = col_character(),.default=col_double())) %>%
    mutate(level="asv")

  model_df <- bind_rows(phylum,class,order,family,genus,otu,asv)

@@

dada2_df <- NULL
dada2_auc_df <- NULL
for( m in models ){
  asv <- read_csv(paste0("../data/dada2/process/combined-dada2-",m,".csv"),
                     col_types = c(method = col_character(),.default=col_double())) %>%
    mutate(type="DADA2")
  dada2_df <- bind_rows(dada2_df,asv)

@@

plot_top_imp <- function(top_data,tax_level){
  top_data %>% 
    filter(level == tax_level) %>% 
    arrange(rank) %>% 
    ggplot(aes(x=AUC_diff,y=reorder(name,desc(rank)))) +
      #geom_jitter(color=colors2[tax_level],fill=alpha(colors2[tax_level],0.6)) +
      stat_summary(fun.data=median_hilow, fun.args=0.5, geom="pointrange", color=colors2[tax_level]) +
      theme_bw() +
      theme(axis.title.y = element_blank(),
        axis.text.y = element_markdown(),
        plot.caption = element_markdown(lineheight = 1.2)) +
      ggtitle(tax_level) +
      xlab("Decrease in AUC")
}

>>>>

library(tidyverse)
library(knitr)
library(kableExtra)
library(ggpubr)
library(nationalparkcolors)
library(data.table)
library(cowplot)
library(pROC)
library(schtools)

@@

read_combined <- function(tax_level,model){
  read_csv(paste0("../data/process/combined-",tax_level,"-",model,".csv"),
           col_types = c(method = col_character(),.default=col_double()),
           na = c("","NA","NaN")) %>% #some of the Pos/Neg Pred Values were NaN
    mutate(level=tax_level) 
}

for( m in models ){
  model_df <- map_dfr(levels, read_combined, model = m)

@@

dada2_df <- NULL
dada2_auc_df <- NULL
for( m in models ){
  asv <- read_csv(paste0("../data/dada2/process/combined-dada2-",m,".csv"),
                  col_types = c(method = col_character(),.default=col_double()),
                  na = c("","NA","NaN")) %>%
    mutate(type="DADA2")
  dada2_df <- bind_rows(dada2_df,asv)

@@

plot_top_imp <- function(top_data,tax_level){
  top_data %>% 
    filter(level == tax_level) %>% 
    arrange(rank) %>% 
    ggplot(aes(x=AUC_diff,y=reorder(name,desc(rank)))) +
      #geom_jitter(color=colors2[tax_level],fill=alpha(colors2[tax_level],0.6)) +
      stat_summary(fun.data=median_hilow, fun.args=0.5, geom="pointrange", color=colors2[tax_level]) +
      # stat_summary(fun = mean,geom = "pointrange",  color=colors2[tax_level],
      #              fun.max = function(x) mean(x) + sd(x),
      #              fun.min = function(x) mean(x) - sd(x)) +
      theme_bw() +
      theme(axis.title.y = element_blank(),
        axis.text.y = element_markdown(),
        plot.caption = element_markdown(lineheight = 1.2)) +
      ggtitle(tax_level) +
      xlab("Decrease in AUC")
}

### 2021-08-30 paper/manuscript.Rmd
draft to send to coauthors

```{r setup_environment, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message=FALSE)
options(scipen=999)
library(tidyverse)
library(kableExtra)
library(schtools)
```

@@

Courtney R. Armour, Begüm D.Topçuoğlu, Andrea Garretto, Patrick D. Schloss ${^\dagger}$

${\dagger}$ To whom correspondence should be addressed:

\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

Department of Microbiology

University of Michigan

Ann Arbor, MI 48109

@@

```{r median_AUC}
table <- read_csv("../exploratory/figures/median_test_auc.csv",
                  col_types = cols(level=col_character(),
                                   .default=col_double()))

rf_med_auc <- table %>% 
  pull(rf)

names(rf_med_auc) <- table %>% pull(level)
```

@@

```{r dada2}
dada2_results <- read_csv("../data/dada2/process/combined-dada2-rf.csv",
                          col_types = cols(method = col_character(),
                                           .default = col_double()))
dada2_med_auc <- dada2_results %>% 
  pull(AUC) %>% 
  median()

@@

Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated abundance tables with mothur {} annotated to phylum, class, order, family, genus, OTU and ASV levels. Using the taxonomic abundance data, we quantified how reliably samples could be classified as "normal" or "SRN" (i.e. advanced adenoma or carcinoma) using five machine learning methods with the mikropml R package (methods). Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1). Regardless of the taxonomic level, random forest (RF) models consistently had the largest area under the receiver operating characteristic curve (AUROC). Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1A, Supplemental Figure 2). Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTUs (p \< 0.01), but comparable to family (p = `r round(rf_pvals["family","asv"],digits=2)`) and genus (p = `r round(rf_pvals["genus","asv"],digits=2)`) levels (Figure 1A). These results suggest that finer taxonomic resolution is not necessarily better for identifying individuals with SRNs based on microbiome composition.

While comparing AUROC values between models is a useful way to assess the overall model performance, they summarize the performance across all thresholds and can be misleading since models with the same AUC can have different ROC curve shapes [@lobo2008]. Depending on the intended implementation of the model, one may want to optimize the true-positive rate (or sensitivity) over the false-positive rate (or 1-specificity), or vice versa. In this case, the optimal model will detect as many true positives (people with SRNs) as possible. To further compare the model performance across taxonomic levels we compared the sensitivity of the models at a specificity of 90%. The highest sensitivity values were observed for family, genus, and OTU level data (Figure 1B), consistent with the AUROC results. phylum, class, order, and ASV sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (Figure 1B). This analysis further supports the observation that finer resolution is does not improve SRN detection.

One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g. phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls. At the level of genus or OTU there is enough data and variation but at the ASV level, the data is too specific to individuals and does not overlap enough. Examination of the prevalence of taxa in samples at each level supports this idea. A majority of taxa were present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels. The opposite was observed at the OTU and ASV level where 60% and 53% of taxa respectively were only present in less than 25% of samples (Supplemental Figure 3). Of note, the ML pipeline includes a pre-processing step that occured prior to training and classifying the ML models. As part of this step, taxa were removed that would not provide useful information to build the model. For example, perfectly correlated taxa provide the same information to build the model and thus can be collapsed. Additionally, features with zero or near-zero variance across samples would not help the model differentiate groups and thus can be removed. Interestingly, despite starting with `r format_number(input %>% filter(level == "asv") %>% select(n_features) %>% pull())` ASVs, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing. At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` of the `r format_number(input %>% filter(level == "otu") %>% select(n_features) %>% pull())` OTUs (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1). While the resolution provided by ASVs is useful in certain contexts {}, these results suggest that the resolution is too fine for use in machine learning classification of SRNs based on microbiome composition.

@@

***Data processing.*** Sequence data was processed with mothur (1.44.3) using the SILVA reference database (v132) to produce count tables for phylum, class, order, family, genus, OTU, and ASV following the Schloss Lab MiSeq SOP described on the mothur website. ASV level data was also produced using DADA2 to ensure consistent results with a different pipeline. Data was processed following the DADA2 pipeline, but setting pool=TRUE to infer ASVs from the whole dataset rather than per sample. The resulting ASV table was subsampled for consistency with the mothur data. These results were consistent for ASVs generated with DADA2 [@callahan2016] (median AUROC: `r round(dada2_med_auc,digits=3)`).

@@

**Figure 1: Random Forest Model Performance.** **A)** Boxplots with points of area under the receiver operating characteristic curve (AUROC) values on the test dataset for 100 seeds predicting SRNs using a Random Forest model. Dashed line denotes AUROC of 0.5 which is equivalent to random classification. Significance between taxonomic levels was quantified by comparing the difference in mean AUROC and is denoted by letters A through E on the right side of the plot; taxonomic levels with the same letter are in the same significance group and are not significantly different from one another. **B)** Strip plot of the sensitivity at a specificity of 90% across the 100 model iterations. Black points denote the median and the lines denote the IQR. The letters W through Z denote the significance groups.

@@

read_csv("table_1.csv") %>%
  select(-n_samples) %>%
  mutate(pct_kept=round((n_features_preproc/n_features)*100,digits=1)) %>%
  mutate(pct_kept=paste0(pct_kept," %"),
         n_features = unlist(lapply(n_features,format_number)),
         level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv" ),
                      labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV"))) %>%
  knitr::kable(format="markdown",booktabs=T,escape=F,align="lrrc",
               linesep="",
               col.names = linebreak(c("Taxonomic Level",
                                 "Number of Features",
                                 "Number of Features \n After Preprocessing",
                                 "Percent of Features Kept \n After Preprocessing"),align="c"))

@@

**Supplemental Figure 1: Model Performance across Taxonomy.** Boxplot of AUROC values on the test dataset for 100 seeds for each model type across all taxonomic levels.

@@

**Supplemental Figure 2: Averaged ROC curves** ROC curves averaged across the 100 iterations of the model. The shaded region represents the standard deviation form the mean.

@@

![](figure_s3.png){height="45%"}\
**Supplemental Figure 3: Prevalence of Taxa in Samples.** Distribution of taxa prevalence in samples at each taxonomic level.

@@

![](figure_s4.png){height="85%"}\
**Supplemental Figure 4: Top 10 important taxa at each taxonomic level.** Summary of the 10 most important taxa for the random forest models at each taxonomic level based on the average decrease in AUC when the feature is permuted.

>>>>

```{r setup_environment, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message=FALSE)
options(scipen=999)
library(tidyverse)
library(kableExtra)
library(schtools)

format_decimal <- function(x,digits){
  format(round(x,digits = digits),nsmall=digits)
}
```

@@

Courtney R. Armour${^1}$, Begüm D.Topçuoğlu, Andrea Garretto${^1}$, Patrick D. Schloss ${^1}$${^*}$

\vspace{5mm}

${^1}$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor

\vspace{5mm}

${^*}$ Correspondence: Patrick D. Schloss \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

@@

```{r mean_AUC}
table <- read_csv("../analysis/mean_AUC_by_model.csv",
                  col_types = cols(level=col_character(),
                                   .default=col_double()))

rf_mean_auc <- table %>% 
  pull(rf)

names(rf_mean_auc) <- table %>% pull(level)
```

@@

```{r dada2}
dada2_results <- read_csv("../data/dada2/process/combined-dada2-rf.csv",
                          col_types = cols(method = col_character(),
                                           .default = col_double()))
dada2_mean_auc <- dada2_results %>% 
  pull(AUC) %>% 
  mean()

@@

Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated abundance tables with mothur {} annotated to phylum, class, order, family, genus, OTU and ASV levels. Using the taxonomic abundance data, we quantified how reliably samples could be classified as "normal" or "SRN" (i.e. advanced adenoma or carcinoma) using five machine learning methods with the mikropml R package (methods). Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1). Regardless of the taxonomic level, random forest (RF) models consistently had the largest area under the receiver operating characteristic curve (AUROC). Within the RF model, the highest AUROCs were observed for family (mean AUROC: `r format_decimal(rf_mean_auc["family"],digits=3)`), genus (mean AUROC: `r format_decimal(rf_mean_auc["genus"],digits=3)`), and OTU (mean AUROC: `r format_decimal(rf_mean_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1A, Supplemental Figure 2). Performance with ASVs (mean AUROC: `r format_decimal(rf_mean_auc["asv"],digits=3)`) was significantly lower than OTUs (p \< 0.01), but comparable to family (p = `r format_decimal(rf_pvals["family","asv"],digits=2)`) and genus (p = `r format_decimal(rf_pvals["genus","asv"],digits=2)`) levels (Figure 1A). These results suggest that increased resolution improves model performance until the OTU level where further taxonomic resolution is not necessarily better for identifying individuals with SRNs based on microbiome composition.

While comparing AUROC values between models is a useful way to assess the overall model performance, they summarize the performance across all thresholds and can be misleading since models with the same AUROC can have different ROC curve shapes [@lobo2008]. Depending on the intended implementation of the model, one may want to optimize the true-positive rate (or sensitivity) over the false-positive rate (or 1-specificity), or vice versa. In this case, the optimal model will detect as many true positives (people with SRNs) as possible. To further compare the model performance across taxonomic levels we compared the sensitivity of the models at a specificity of 90%. The highest sensitivity values were observed for family, genus, and OTU level data (Figure 1B), consistent with the AUROC results. Phylum, class, order, and ASV sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (Figure 1B). This analysis further supports the observation that finer resolution is does not improve SRN detection.

```{r prevalence}

bin_prevalence <- function(prev,tax_level){
  sub_data <- prev %>% 
    filter(level==tax_level,kept==1)
  hist <- hist(sub_data$total_prevalence,breaks = seq(0,1,0.1),plot=FALSE)

  nOTU <- sub_data %>%
    select(name) %>% 
    n_distinct()

  tibble(level=tax_level,
         prev=hist$breaks[-1],
         count=hist$counts,
         pct=hist$counts/nOTU)
}

prev <- read_csv("../analysis/prevalence_by_level.csv")

levels <- c("phylum","class","order","family","genus","otu","asv")

pct_bins <- map_dfr(levels,bin_prevalence,prev=prev)

#percent of taxa present in > 70% samples
pct_above <- pct_bins %>% 
  mutate(prev = round(prev,digits = 1)) %>% 
  filter(prev > 0.7) %>% 
  group_by(level)  %>% 
  summarise(prev_above = round((sum(pct)) * 100)) %>% 
  pivot_wider(names_from = level,values_from = prev_above)

#percent of taxa present in < 20% samples
pct_below <- pct_bins %>% 
  mutate(prev = round(prev,digits = 1)) %>% 
  filter(prev <= 0.2) %>% 
  group_by(level)  %>% 
  summarise(prev_above = round((sum(pct)) * 100)) %>% 
  pivot_wider(names_from = level,values_from = prev_above)
```

One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g. phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls. At the level of genus or OTU there is enough data and variation but at the ASV level, the data is too specific to individuals and does not overlap enough. Examination of the prevalence of taxa in samples at each level supports this idea. A majority of taxa were present in greater than 70% of samples at the phylum (`r pct_above$phylum`% of taxa) and class (`r pct_above$class`% of taxa) levels. The opposite was observed at the OTU and ASV level where `r pct_below$otu`% and `r pct_below$asv`% of taxa respectively were only present in 20% or less of the samples (Supplemental Figure 3). Of note, the ML pipeline includes a pre-processing step that occurred prior to training and classifying the ML models. As part of this step, taxa were removed that would not provide useful information to build the model. For example, perfectly correlated taxa provide the same information to build the model and thus can be collapsed. Additionally, features with zero or near-zero variance across samples would not help the model differentiate groups and thus can be removed. Interestingly, despite starting with `r format_number(input %>% filter(level == "asv") %>% select(n_features) %>% pull())` ASVs, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing. At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` of the `r format_number(input %>% filter(level == "otu") %>% select(n_features) %>% pull())` OTUs (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1). While the resolution provided by ASVs is useful in certain contexts {}, these results suggest that the resolution is too fine for use in machine learning classification of SRNs based on microbiome composition.

@@

***Data processing.*** Sequence data was processed with mothur (1.44.3) using the SILVA reference database (v132) to produce count tables for phylum, class, order, family, genus, OTU, and ASV following the Schloss Lab MiSeq SOP described on the mothur website. ASV level data was also produced using DADA2 to ensure consistent results with a different pipeline. Data was processed following the DADA2 pipeline, but setting pool=TRUE to infer ASVs from the whole dataset rather than per sample. The resulting ASV table was subsampled for consistency with the mothur data. These results were consistent for ASVs generated with DADA2 [@callahan2016] (mean AUROC: `r format_decimal(dada2_mean_auc,digits=3)`).

@@

**Figure 1: Random Forest Model Performance.** **A)** Strip plot of the area under the receiver operating characteristic curve (AUROC) values on the test dataset for 100 seeds predicting SRNs using a random forest model. Black points denote the mean and lines denote the standard deviation. Dashed line denotes AUROC of 0.5 which is equivalent to random classification. Significance between taxonomic levels was quantified by comparing the difference in mean AUROC and is denoted by letters A through E on the right side of the plot; taxonomic levels with the same letter are in the same significance group and are not significantly different from one another. **B)** Strip plot of the sensitivity at a specificity of 90% across the 100 model iterations for each taxonomic level. Black points denote the mean and the lines denote the standard deviation. The letters W through Z on the right side of the plot denote the significance groups.

@@

read_csv("table_1.csv") %>%
  select(-n_samples) %>%
  mutate(pct_kept=format_decimal((n_features_preproc/n_features)*100,digits=1)) %>%
  mutate(pct_kept=paste0(pct_kept," %"),
         n_features = unlist(lapply(n_features,format_number)),
         level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv" ),
                      labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV"))) %>%
  knitr::kable(format="markdown",booktabs=T,escape=F,align="lrrc",
               linesep="",
               col.names = linebreak(c("Taxonomic Level",
                                 "Number of Features",
                                 "Number of Features \n After Preprocessing",
                                 "Percent of Features Kept \n After Preprocessing"),align="c"))

@@

**Supplemental Figure 1: Model Performance across Taxonomy.** Boxplots of AUROC values on the test dataset split for 100 seeds for each model type across all taxonomic levels.

@@

**Supplemental Figure 2: Averaged ROC curves** ROC curves with averaged true positive rate across the 100 iterations of the model. The shaded region represents the standard deviation form the mean.Dashed line represents an AUROC of 0.5, which is equivalent to random classification. The mean AUROC for each taxonomic level is printed on the bottom right of the plot.

@@

![](figure_s3.png){height="45%"}\
**Supplemental Figure 3: Prevalence of Taxa in Samples.** Distribution of the prevalence of taxa across samples at each taxonomic level. Percent of samples is split into 10 groups where the first is taxa present in 0 to 10% of samples, then >10% to 20% of samples, and so on. The total number of taxa for each taxonomic level after preprocessing is in parenthesis next to the title of the plot (the name of the taxonomic level).

@@

![](figure_s4.png){height="85%"}\
**Supplemental Figure 4: Top 10 important taxa at each taxonomic level.** Summary of the 10 most important taxa for the random forest models at each taxonomic level based on the average decrease in AUROC when the feature is permuted.

### 2021-08-30 paper/manuscript.tex
draft to send to coauthors

Courtney R. Armour, Begüm D.Topçuoğlu, Andrea Garretto, Patrick D.
Schloss \({^\dagger}\)

\({\dagger}\) To whom correspondence should be addressed:

\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

Department of Microbiology

University of Michigan

Ann Arbor, MI 48109

@@

Utilizing publicly available 16S rRNA sequence data from stool of
patients with SRNs and healthy controls, we generated abundance tables
with mothur \{\} annotated to phylum, class, order, family, genus, OTU
and ASV levels. Using the taxonomic abundance data, we quantified how
reliably samples could be classified as ``normal'' or ``SRN''
(i.e.~advanced adenoma or carcinoma) using five machine learning methods
with the mikropml R package (methods). Across the five machine learning
methods tested, model performance tended to increase with taxonomic
level usually peaking around genus/OTU level before dropping off
slightly with ASVs (Supplemental Figure 1). Regardless of the taxonomic
level, random forest (RF) models consistently had the largest area under
the receiver operating characteristic curve (AUROC). Within the RF
model, the highest AUROCs were observed for family (median AUROC:
0.687), genus (median AUROC: 0.686), and OTU (median AUROC: 0.698) level
data with no significant difference between the three (Figure 1A,
Supplemental Figure 2). Performance with ASVs (median AUROC: 0.676) was
significantly lower than OTUs (p \textless{} 0.01), but comparable to
family (p = 0.06) and genus (p = 0.05) levels (Figure 1A). These results
suggest that finer taxonomic resolution is not necessarily better for
identifying individuals with SRNs based on microbiome composition.

While comparing AUROC values between models is a useful way to assess
the overall model performance, they summarize the performance across all
thresholds and can be misleading since models with the same AUC can have
different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{5}).
Depending on the intended implementation of the model, one may want to
optimize the true-positive rate (or sensitivity) over the false-positive
rate (or 1-specificity), or vice versa. In this case, the optimal model
will detect as many true positives (people with SRNs) as possible. To
further compare the model performance across taxonomic levels we
compared the sensitivity of the models at a specificity of 90\%. The
highest sensitivity values were observed for family, genus, and OTU
level data (Figure 1B), consistent with the AUROC results. phylum,
class, order, and ASV sensitivity values were all significantly lower
than family, genus, and OTU sensitivity values (Figure 1B). This
analysis further supports the observation that finer resolution is does
not improve SRN detection.

One hypothesis for the observation that model performance increases from
phylum to OTU level then drops slightly at ASV level is that at higher
taxonomic levels (e.g.~phylum) there are too few taxa and too much
overlap to reliably differentiate between cases and controls. At the
level of genus or OTU there is enough data and variation but at the ASV
level, the data is too specific to individuals and does not overlap
enough. Examination of the prevalence of taxa in samples at each level
supports this idea. A majority of taxa were present in greater than 75\%
of samples at the phylum (67\% of taxa) and class (63\% of taxa) levels.
The opposite was observed at the OTU and ASV level where 60\% and 53\%
of taxa respectively were only present in less than 25\% of samples
(Supplemental Figure 3). Of note, the ML pipeline includes a
pre-processing step that occured prior to training and classifying the
ML models. As part of this step, taxa were removed that would not
provide useful information to build the model. For example, perfectly
correlated taxa provide the same information to build the model and thus
can be collapsed. Additionally, features with zero or near-zero variance
across samples would not help the model differentiate groups and thus
can be removed. Interestingly, despite starting with 104,106 ASVs, only
478 (0.5\%) remained after pre-processing. At the OTU level, 705 of the
20,079 OTUs (3.5\%) remained after preprocessing (Table 1). While the
resolution provided by ASVs is useful in certain contexts \{\}, these
results suggest that the resolution is too fine for use in machine
learning classification of SRNs based on microbiome composition.

@@

\textbf{\emph{Data processing.}} Sequence data was processed with mothur
(1.44.3) using the SILVA reference database (v132) to produce count
tables for phylum, class, order, family, genus, OTU, and ASV following
the Schloss Lab MiSeq SOP described on the mothur website. ASV level
data was also produced using DADA2 to ensure consistent results with a
different pipeline. Data was processed following the DADA2 pipeline, but
setting pool=TRUE to infer ASVs from the whole dataset rather than per
sample. The resulting ASV table was subsampled for consistency with the
mothur data. These results were consistent for ASVs generated with DADA2
(\protect\hyperlink{ref-callahan2016}{6}) (median AUROC: 0.66).

@@

\textbf{Figure 1: Random Forest Model Performance.} \textbf{A)} Boxplots
with points of area under the receiver operating characteristic curve
(AUROC) values on the test dataset for 100 seeds predicting SRNs using a
Random Forest model. Dashed line denotes AUROC of 0.5 which is
equivalent to random classification. Significance between taxonomic
levels was quantified by comparing the difference in mean AUROC and is
denoted by letters A through E on the right side of the plot; taxonomic
levels with the same letter are in the same significance group and are
not significantly different from one another. \textbf{B)} Strip plot of
the sensitivity at a specificity of 90\% across the 100 model
iterations. Black points denote the median and the lines denote the IQR.
The letters W through Z denote the significance groups.

@@

\textbf{Supplemental Figure 1: Model Performance across Taxonomy.}
Boxplot of AUROC values on the test dataset for 100 seeds for each model
type across all taxonomic levels.

@@

\textbf{Supplemental Figure 2: Averaged ROC curves} ROC curves averaged
across the 100 iterations of the model. The shaded region represents the
standard deviation form the mean.

@@

\includegraphics[width=\textwidth,height=0.45\textheight]{figure_s3.png}\\
\textbf{Supplemental Figure 3: Prevalence of Taxa in Samples.}
Distribution of taxa prevalence in samples at each taxonomic level.

@@

\includegraphics[width=\textwidth,height=0.85\textheight]{figure_s4.png}\\
\textbf{Supplemental Figure 4: Top 10 important taxa at each taxonomic
level.} Summary of the 10 most important taxa for the random forest
models at each taxonomic level based on the average decrease in AUC when
the feature is permuted.

>>>>

Courtney R. Armour\({^1}\), Begüm D.Topçuoğlu, Andrea Garretto\({^1}\),
Patrick D. Schloss \({^1}\)\({^*}\)

\vspace{5mm}

\({^1}\) Department of Microbiology and Immunology, University of
Michigan, Ann Arbor

\vspace{5mm}

\({^*}\) Correspondence: Patrick D. Schloss
\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

@@

Utilizing publicly available 16S rRNA sequence data from stool of
patients with SRNs and healthy controls, we generated abundance tables
with mothur \{\} annotated to phylum, class, order, family, genus, OTU
and ASV levels. Using the taxonomic abundance data, we quantified how
reliably samples could be classified as ``normal'' or ``SRN''
(i.e.~advanced adenoma or carcinoma) using five machine learning methods
with the mikropml R package (methods). Across the five machine learning
methods tested, model performance tended to increase with taxonomic
level usually peaking around genus/OTU level before dropping off
slightly with ASVs (Supplemental Figure 1). Regardless of the taxonomic
level, random forest (RF) models consistently had the largest area under
the receiver operating characteristic curve (AUROC). Within the RF
model, the highest AUROCs were observed for family (mean AUROC: 0.689),
genus (mean AUROC: 0.690), and OTU (mean AUROC: 0.693) level data with
no significant difference between the three (Figure 1A, Supplemental
Figure 2). Performance with ASVs (mean AUROC: 0.676) was significantly
lower than OTUs (p \textless{} 0.01), but comparable to family (p =
0.06) and genus (p = 0.05) levels (Figure 1A). These results suggest
that increased resolution improves model performance until the OTU level
where further taxonomic resolution is not necessarily better for
identifying individuals with SRNs based on microbiome composition.

While comparing AUROC values between models is a useful way to assess
the overall model performance, they summarize the performance across all
thresholds and can be misleading since models with the same AUROC can
have different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{5}).
Depending on the intended implementation of the model, one may want to
optimize the true-positive rate (or sensitivity) over the false-positive
rate (or 1-specificity), or vice versa. In this case, the optimal model
will detect as many true positives (people with SRNs) as possible. To
further compare the model performance across taxonomic levels we
compared the sensitivity of the models at a specificity of 90\%. The
highest sensitivity values were observed for family, genus, and OTU
level data (Figure 1B), consistent with the AUROC results. Phylum,
class, order, and ASV sensitivity values were all significantly lower
than family, genus, and OTU sensitivity values (Figure 1B). This
analysis further supports the observation that finer resolution is does
not improve SRN detection.

One hypothesis for the observation that model performance increases from
phylum to OTU level then drops slightly at ASV level is that at higher
taxonomic levels (e.g.~phylum) there are too few taxa and too much
overlap to reliably differentiate between cases and controls. At the
level of genus or OTU there is enough data and variation but at the ASV
level, the data is too specific to individuals and does not overlap
enough. Examination of the prevalence of taxa in samples at each level
supports this idea. A majority of taxa were present in greater than 70\%
of samples at the phylum (67\% of taxa) and class (63\% of taxa) levels.
The opposite was observed at the OTU and ASV level where 50\% and 41\%
of taxa respectively were only present in 20\% or less of the samples
(Supplemental Figure 3). Of note, the ML pipeline includes a
pre-processing step that occurred prior to training and classifying the
ML models. As part of this step, taxa were removed that would not
provide useful information to build the model. For example, perfectly
correlated taxa provide the same information to build the model and thus
can be collapsed. Additionally, features with zero or near-zero variance
across samples would not help the model differentiate groups and thus
can be removed. Interestingly, despite starting with 104,106 ASVs, only
478 (0.5\%) remained after pre-processing. At the OTU level, 705 of the
20,079 OTUs (3.5\%) remained after preprocessing (Table 1). While the
resolution provided by ASVs is useful in certain contexts \{\}, these
results suggest that the resolution is too fine for use in machine
learning classification of SRNs based on microbiome composition.

@@

\textbf{\emph{Data processing.}} Sequence data was processed with mothur
(1.44.3) using the SILVA reference database (v132) to produce count
tables for phylum, class, order, family, genus, OTU, and ASV following
the Schloss Lab MiSeq SOP described on the mothur website. ASV level
data was also produced using DADA2 to ensure consistent results with a
different pipeline. Data was processed following the DADA2 pipeline, but
setting pool=TRUE to infer ASVs from the whole dataset rather than per
sample. The resulting ASV table was subsampled for consistency with the
mothur data. These results were consistent for ASVs generated with DADA2
(\protect\hyperlink{ref-callahan2016}{6}) (mean AUROC: 0.659).

@@

\textbf{Figure 1: Random Forest Model Performance.} \textbf{A)} Strip
plot of the area under the receiver operating characteristic curve
(AUROC) values on the test dataset for 100 seeds predicting SRNs using a
random forest model. Black points denote the mean and lines denote the
standard deviation. Dashed line denotes AUROC of 0.5 which is equivalent
to random classification. Significance between taxonomic levels was
quantified by comparing the difference in mean AUROC and is denoted by
letters A through E on the right side of the plot; taxonomic levels with
the same letter are in the same significance group and are not
significantly different from one another. \textbf{B)} Strip plot of the
sensitivity at a specificity of 90\% across the 100 model iterations for
each taxonomic level. Black points denote the mean and the lines denote
the standard deviation. The letters W through Z on the right side of the
plot denote the significance groups.

@@

\textbf{Supplemental Figure 1: Model Performance across Taxonomy.}
Boxplots of AUROC values on the test dataset split for 100 seeds for
each model type across all taxonomic levels.

@@

\textbf{Supplemental Figure 2: Averaged ROC curves} ROC curves with
averaged true positive rate across the 100 iterations of the model. The
shaded region represents the standard deviation form the mean.Dashed
line represents an AUROC of 0.5, which is equivalent to random
classification. The mean AUROC for each taxonomic level is printed on
the bottom right of the plot.

@@

\includegraphics[width=\textwidth,height=0.45\textheight]{figure_s3.png}\\
\textbf{Supplemental Figure 3: Prevalence of Taxa in Samples.}
Distribution of the prevalence of taxa across samples at each taxonomic
level. Percent of samples is split into 10 groups where the first is
taxa present in 0 to 10\% of samples, then \textgreater10\% to 20\% of
samples, and so on. The total number of taxa for each taxonomic level
after preprocessing is in parenthesis next to the title of the plot (the
name of the taxonomic level).

@@

\includegraphics[width=\textwidth,height=0.85\textheight]{figure_s4.png}\\
\textbf{Supplemental Figure 4: Top 10 important taxa at each taxonomic
level.} Summary of the 10 most important taxa for the random forest
models at each taxonomic level based on the average decrease in AUROC
when the feature is permuted.
